---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.path = "man/figures/README-", out.width = "100%")
pkgload::load_all(".", quiet = TRUE)
```

# chorale

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/lescailab/chorale/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/lescailab/chorale/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/lescailab/chorale/graph/badge.svg)](https://app.codecov.io/gh/lescailab/chorale)
<!-- badges: end -->

CHORALE analyses omics modalities measured on different individuals by first
placing every sample in the same space of named biological concepts.

The modalities share no sample and need share no feature. Their connection is a
fixed vocabulary: pathways, cell-type signatures, or other feature sets chosen
before the results are inspected. Once each modality has been scored on that
vocabulary, CHORALE supports two complementary analyses.

- **Concept evidence** estimates the adjusted phenotype effect separately in
  each modality and combines those effects by inverse-variance weighting.
- **Joint-state analysis** stacks all samples by their concept scores and fits a
  low-rank representation, then tests and transfers the resulting directions.

Related concepts can also be tested as families, while variation outside the
vocabulary is retained as free dimensions.

The shared preparation runs in a fixed order:

1. The vocabulary is fixed and its coverage reported, per concept and per modality.
2. Every modality is scored on it, without the design being consulted.
3. What the vocabulary does not explain is kept as free dimensions, selected by reproducibility.
4. The concept branch regresses each score on the design within each modality
   and combines the adjusted effects.
5. The joint branch standardises scores within modality, optionally removes
   modality-local nuisance effects, and fits the observed entries of the
   stacked score matrix.
6. Freedman--Lane residual permutations hold the design fixed while calibrating
   concept, family, joint-component, and free-dimension evidence.

## What a result says

Each concept carries its adjusted phenotype effect in every modality that
expresses it, the combined statistic, agreement diagnostics, family-wise and
false-discovery control, and the statistic that survives adjustment for
overlapping neighbours. A high heterogeneity p-value means that disagreement
was not detected; it does not prove that the effects are identical.

A joint component is a fitted direction over concepts, not an identified
mechanism. Its sign and label are arbitrary, so it is interpreted through its
loadings, calibrated phenotype evidence, per-modality effects, and
leave-one-modality-out transfer. Coordinated variation no concept explains is
reported separately as free dimensions.

## Status

This branch is a production candidate for external review, not a stable or CRAN release. Validation
includes planted-concept simulations, complete-null calibration, recovery measured against a
withheld pairing, deterministic tests, package checks, and site builds. Conditions of applicability
are stated in the [methods documentation](https://lescailab.github.io/chorale/methods).

## Installation

```r
# install.packages("devtools")
devtools::install_github("lescailab/chorale")
```

For a reproducible development environment:

```bash
conda env create -f environment.yml
conda activate chorale_development
```

The package requires R 4.5. The repository-owned environment uses R 4.5 and Node 22 and lists direct dependencies without
machine-specific build hashes.

## Minimal example

```{r example}
library(chorale)

# Two cohorts on disjoint individuals, with one named concept planted in both.
fx <- chorale_concept_example(seed = 1)

fit <- chorale_concept_fit(fx$containers, fx$sets, n_free = 1,
                           n_permutations = 199, n_init = 3)
fit$evidence$joint[, c("concept", "n_modalities", "joint_z",
                       "sign_agreement", "q_value")]

# The same encoding can be analysed jointly. This does not happen implicitly:
# component count and nuisance adjustment remain explicit decisions.
state <- chorale_joint_state(fit$encoding, n_components = 1)
chorale_joint_evidence(state, n_permutations = 199)$components
```

Use `chorale_control()` to set the phenotype column, reference level, exchangeability blocks, the
reproducibility threshold the free dimensions are selected at, and the error rate concepts are
called at.

## Documentation

The repository vignettes are the canonical source for the
[documentation site](https://lescailab.github.io/chorale):

- [Why CHORALE exists](https://lescailab.github.io/chorale/why)
- [How the estimator works](https://lescailab.github.io/chorale/how-it-works)
- [The input data](https://lescailab.github.io/chorale/input-format)
- [Tutorial](https://lescailab.github.io/chorale/tutorial)
- [Outputs and how to read them](https://lescailab.github.io/chorale/outputs)
- [Statistical methods](https://lescailab.github.io/chorale/methods)
- [Simulation and validation](https://lescailab.github.io/chorale/simulation)
