---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      fig.path = "man/figures/README-", out.width = "100%")
```

# chorale

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/lescailab/chorale/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/lescailab/chorale/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/lescailab/chorale/graph/badge.svg)](https://app.codecov.io/gh/lescailab/chorale)
<!-- badges: end -->

CHORALE estimates a shared biological state across omics modalities measured on different
individuals, and expresses it in named biological terms.

The modalities share no sample and need share no feature, so what connects them is a fixed
vocabulary of named biological concepts. A concept is a set of features that belong together for a
stated biological reason, such as a curated pathway or a cell type signature. Every modality is
scored on the same concepts, a concept is the same concept in every modality, and the design then
supplies the evidence that a concept separates cases from controls.

The estimator runs in a fixed order:

1. The vocabulary is fixed and its coverage reported, per concept and per modality.
2. Every modality is scored on it, without the design being consulted.
3. What the vocabulary does not explain is kept as free dimensions, selected by reproducibility.
4. The concept score is regressed on the design in each modality, and the modalities are combined by
   inverse-variance weighting.
5. Permuting the phenotype within strata of the design calibrates the whole vocabulary at once.

## What a result says

Each concept carries its adjusted case-control effect in every modality that expresses it, the
combined statistic, how far the modalities agree in direction and in magnitude, family-wise and
false-discovery control over the vocabulary, and the statistic that survives once overlapping
concepts are regressed out. Coordinated variation no concept explains is reported separately as
free dimensions, so the bound the vocabulary places on interpretation stays visible.

A supported concept is a coordinated case-control shift in a named piece of biology, seen in more
than one measurement layer, at a stated error rate. Ordering cause and consequence requires timing,
mediation or perturbation.

## Status

This branch is a production candidate for external review, not a stable or CRAN release. Validation
includes planted-concept simulations, complete-null calibration, recovery measured against a
withheld pairing, deterministic tests, package checks, and site builds. Conditions of applicability
are stated in the [methods documentation](https://lescailab.github.io/chorale/methods).

## Installation

```r
# install.packages("devtools")
devtools::install_github("lescailab/chorale")
```

For a reproducible development environment:

```bash
conda env create -f environment.yml
conda activate chorale_development
```

The repository-owned environment uses R 4.5 and Node 22 and lists direct dependencies without
machine-specific build hashes.

## Minimal example

```{r example}
library(chorale)

# Two cohorts on disjoint individuals, with one named concept planted in both.
fx <- chorale_concept_example(seed = 1)

fit <- chorale_concept_fit(fx$containers, fx$sets, n_free = 1,
                           n_permutations = 199, n_init = 3)
fit$evidence$joint[, c("concept", "n_modalities", "joint_z",
                       "sign_agreement", "q_value")]
```

Use `chorale_control()` to set the phenotype column, reference level, exchangeability blocks, the
reproducibility threshold the free dimensions are selected at, and the error rate concepts are
called at.

## Documentation

The repository vignettes are the canonical source for the
[documentation site](https://lescailab.github.io/chorale):

- [Why CHORALE exists](https://lescailab.github.io/chorale/why)
- [How the estimator works](https://lescailab.github.io/chorale/how-it-works)
- [The input data](https://lescailab.github.io/chorale/input-format)
- [Tutorial](https://lescailab.github.io/chorale/tutorial)
- [Outputs and how to read them](https://lescailab.github.io/chorale/outputs)
- [Statistical methods](https://lescailab.github.io/chorale/methods)
- [Simulation and validation](https://lescailab.github.io/chorale/simulation)
