---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# chorale

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/lescailab/chorale/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/lescailab/chorale/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/lescailab/chorale/graph/badge.svg)](https://app.codecov.io/gh/lescailab/chorale)
<!-- badges: end -->

chorale estimates a shared latent biological state across omics modalities
measured on **disjoint sets of animals**, where no subject contributes to more
than one modality, and expresses that state in biological terms.

The framework rests on a single result: in unpaired multi-domain linear latent
models the shared latent block is identifiable from marginal distributions
alone, because the modality-specific components are non-exchangeable and
non-Gaussian (Sturma et al., NeurIPS 2023, Theorem 3.1; Timilsina, Lyu & Fu,
arXiv:2409.19422, Theorem 1). Identification fails where those modality-specific
structures are made statistically alike (Sturma et al., Proposition C.1 and
Lemma C.2). Modality heterogeneity is therefore treated as the identifying
resource, and cross-modality harmonisation is excluded.

Two properties hold throughout.

**Interpretability by design.** Every latent factor carries a biological
definition from the moment it is estimated, through pathway-constrained loadings
and a pure-feature requirement, not by post-hoc annotation.

**Falsifiability.** Every result is accompanied by its negative control:
permutation calibration, modality shuffle, and stability across random
initialisations. No output is reportable without one.

## Status

Under development. Containers, identifier mapping and the synthetic data
generator are implemented; the estimator is scaffolded but not yet implemented,
pending the feasibility gates that determine whether the linear non-Gaussian
identification route applies to the data at hand.

| Function | Status |
|---|---|
| `chorale_load()` | Implemented |
| `chorale_map()` | Implemented |
| `chorale_simulate()` | Implemented |
| `chorale_python_setup()` | Implemented |
| `chorale_fit()` | Scaffolded |
| `chorale_bound()` | Scaffolded |
| `chorale_null()` | Scaffolded |
| `chorale_report()` | Scaffolded |

## Installation

```r
# install.packages("devtools")
devtools::install_github("lescailab/chorale")
```

Bioconductor dependencies resolve through `BiocManager`. Comparator methods
listed in `Suggests` are optional: their absence degrades a comparison rather
than breaking the package.

## Example

Generate disjoint cohorts over a shared feature space with known ground truth,
then load one modality into the common container:

```{r example}
library(chorale)

sim <- chorale_simulate(
  n_modalities = 3,
  n_features = 200,
  n_shared_factors = 4,
  n_strains = 6,
  seed = 1
)

# No animal appears in more than one modality.
length(intersect(sim$col_data[[1]]$sample_id, sim$col_data[[2]]$sample_id))

se <- chorale_load(sim$modalities[[1]], sim$col_data[[1]])
se
```

Harmonise feature identifiers to mouse Entrez, keeping one-to-many mappings with
fractional weight rather than resolving them arbitrarily:

```{r map}
chorale_map(c("Bdnf", "Trem2", "App"), from = "SYMBOL")
```

## Scope

The estimand is the conditional law of the latent state given the design
covariates, together with the modality-specific measurement operators.
Individual-level cross-modality inference is undefined, because each animal is
measured in exactly one modality. The joint distribution across modalities is
not identified nonparametrically, so partial-identification bounds on the
cross-modality coupling are reported alongside any point estimate.

## References

- Sturma N, Squires C, Drton M, Uhler C. Unpaired Multi-Domain Causal
  Representation Learning. *NeurIPS* 2023.
- Timilsina S, Lyu Q, Fu X et al. Identifiable Shared Component Analysis of
  Unpaired Multimodal Mixtures. arXiv:2409.19422.
- Mao W, Zaslavsky E, Hartmann BM, Sealfon SC, Chikina M. Pathway-level
  information extractor (PLIER) for gene expression data. *Nat Methods*
  16:607-610, 2019.
- Taroni JN, Grayson PC, Hu Q et al. MultiPLIER: a transfer learning framework
  for transcriptomics reveals systemic features of rare disease. *Cell Systems*
  8(5):380-394, 2019.
