---
title: "Tutorial"
description: "One complete analysis: concepts, families, joint state, transfer, free dimensions, controls, and report files."
site_order: 4
site_wide: true
output:
  rmarkdown::html_vignette:
    math_method: mathml
vignette: >
  %\VignetteIndexEntry{Tutorial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This tutorial builds three cohorts with no individual in common. Two overlapping
concepts move together, one other concept moves in the opposite direction, one
modality carries a local batch effect, and the assays retain variation outside
the vocabulary. The same fitted encoding is then read in five ways.

The example uses 99 permutations so the vignette builds quickly. Its smallest
attainable p-value is therefore 0.01. A real analysis should choose the count
from the resolution it needs before fitting; 999 or 9,999 are common starting
points for thresholds near 0.05 or 0.01.

## Install and load

```r
# install.packages("devtools")
devtools::install_github("lescailab/chorale")
```

```{r tutorial-setup}
library(chorale)
set.seed(11)
```

## 1. Build three disjoint cohorts

The feature names and concept labels are generic. In a real analysis they would
come from the assay annotation and a curated or custom vocabulary.

```{r tutorial-data}
n <- 48L
p <- 260L
ids <- sprintf("feature_%03d", seq_len(p))

sets <- list(
  process_core      = ids[1:50],
  process_extension = ids[31:80],
  transport         = ids[91:140],
  immune            = ids[151:200],
  quiet             = ids[181:230]
)

make_modality <- function(label, local_batch = FALSE) {
  sample_id <- sprintf("%s_%03d", label, seq_len(n))
  phenotype <- rep(c("reference", "target"), length.out = n)
  sex <- rep(c("F", "F", "M", "M"), length.out = n)

  shared <- ifelse(phenotype == "target", 1, -1) + rnorm(n, sd = 0.65)
  unnamed <- ifelse(phenotype == "target", 0.6, -0.6) + rnorm(n, sd = 0.8)
  x <- matrix(rnorm(p * n), p, dimnames = list(ids, sample_id))
  x[1:50, ] <- x[1:50, ] + 0.38 * rep(shared, each = 50)
  x[31:80, ] <- x[31:80, ] + 0.32 * rep(shared, each = 50)
  x[91:140, ] <- x[91:140, ] - 0.30 * rep(shared, each = 50)
  x[231:260, ] <- x[231:260, ] + 0.65 * rep(unnamed, each = 30)

  design <- data.frame(sample_id, phenotype, sex,
                       stringsAsFactors = FALSE)
  if (local_batch) {
    design$local_batch <- rep(c("b1", "b2"), each = n / 2)
    x[1:80, design$local_batch == "b2"] <-
      x[1:80, design$local_batch == "b2"] + 1.2
  }
  chorale_load(x, design)
}

containers <- list(
  modality_a = make_modality("a", local_batch = TRUE),
  modality_b = make_modality("b"),
  modality_c = make_modality("c")
)

length(Reduce(intersect, lapply(containers, colnames)))
```

The result is zero: no sample appears in two modalities.

## 2. Declare the analysis decisions and coverage

```{r tutorial-concepts}
control <- chorale_control(
  phenotype_reference = "reference",
  n_select_init = 2,
  n_subsample = 2
)

concepts <- chorale_concepts(containers, sets, min_features = 5)
concepts$coverage
```

Coverage answers whether the question can be asked: it reports how far each
modality's annotation reaches into a concept, rather than whether a process is
active. Every synthetic modality expresses every concept here.

## 3. Fit the common encoding and concept evidence

```{r tutorial-fit}
fit <- chorale_concept_fit(
  containers, concepts,
  n_free = 1,
  n_permutations = 99,
  n_init = 2,
  control = control,
  seed = 11
)

fit$evidence$joint[, c(
  "concept", "n_modalities", "joint_z", "sign_agreement",
  "heterogeneity_p", "p_family", "q_value"
)]
```

`process_core`, `process_extension`, and `transport` have large combined
statistics. The sign of `transport` is negative because its member features
were shifted in the opposite direction. `sign_agreement = 1` means all three
modalities point with the combination.

Read effect size from the per-modality table, not from the z-statistic:

```{r tutorial-effects}
fit$evidence$per_modality[
  fit$evidence$per_modality$concept == "process_core",
  c("concept", "modality", "n_features", "effect", "se", "z")
]
```

The effect is measured in standard deviations of the concept score after
adjustment. The z-statistic divides that effect by its uncertainty and therefore
grows with precision. A large `heterogeneity_p` means this test detected no
excess disagreement, which across three modalities it has little power to do.

## 4. Test overlapping concepts as families

```{r tutorial-families}
membership <- chorale_concept_families(concepts, min_overlap = 0.2)
family_evidence <- chorale_family_evidence(
  fit$evidence, membership, control = control
)
family_evidence$families[, c(
  "family", "n_concepts", "statistic", "best_single_z",
  "p_family", "q_value", "members"
)]
```

`process_core` and `process_extension` form one supported family. Their signed
statistics accumulate because they point the same way. `immune` and `quiet`
also overlap, but their family does not beat the family-wise null. Family
significance is a group-level result, and each member is read from the concept
table on its own.

## 5. Estimate and test the joint state

The local batch is recorded only in `modality_a`. It cannot enter a model of
covariates shared by all modalities, but it can be removed from that modality's
scores before stacking.

```{r tutorial-joint}
state <- chorale_joint_state(
  fit$encoding,
  n_components = 2,
  nuisance = list(modality_a = "local_batch"),
  control = control,
  seed = 11
)

joint_evidence <- chorale_joint_evidence(
  state, n_permutations = 99, control = control, seed = 11
)
joint_evidence$components
```

`joint_01` carries the multiconcept phenotype direction. Its effect sign is not
intrinsic: multiplying both its scores and loadings by -1 gives the same fitted
state. Interpret its magnitude and p-values, then inspect its loadings:

```{r tutorial-loadings}
state$loadings[order(abs(state$loadings[, "joint_01"]), decreasing = TRUE),
               "joint_01", drop = FALSE]
joint_evidence$per_modality
```

Per-modality effects show where the fitted component is carried. They are
descriptive context and do not enter the joint p-value.

## 6. Transfer the direction to held-out modalities

```{r tutorial-transfer}
joint_transfer <- chorale_joint_transfer(
  fit$encoding,
  n_components = 2,
  n_permutations = 99,
  nuisance = list(modality_a = "local_batch"),
  control = control,
  seed = 11
)

joint_transfer$transfer[, c(
  "held_out", "trained_on", "component", "fitted_component",
  "loading_agreement", "z", "p_value"
)]
```

`joint_01` transfers to all three held-out modalities at the smallest p-value
this example can attain. `component` is the full-data reference label;
`fitted_component` is its name inside that held-out fit. A negative
`loading_agreement` indicates reversed orientation. Transfer `p_value` is
component-specific; correction across the transfer table is not applied.

## 7. Read variation outside the vocabulary

```{r tutorial-free}
free <- chorale_free_dimensions(fit, n_permutations = 99)
free[, c("modality", "dimension", "variance_share", "reproducibility",
         "z", "p_family", "outside_vocabulary", "top_features")]
```

These dimensions come from the assay residual after projecting onto the whole
concept-score span. In this particular draw none clears the family-wise
threshold. That is a tested negative result, not an omitted analysis. If the
concept-score matrix had spanned the entire centred sample space, no free
dimension would have been estimable; `variance.tsv` distinguishes that case.

## 8. Run controls and write the complete report

```{r tutorial-report}
controls <- chorale_null(fit, containers, n_shuffles = 0)

out_dir <- file.path(tempdir(), "chorale_tutorial_report")
written <- chorale_report(
  fit, null = controls, path = out_dir,
  family_evidence = family_evidence,
  joint_state = state,
  joint_evidence = joint_evidence,
  joint_transfer = joint_transfer
)
basename(written)
```

The report writes only optional analyses supplied to it. It never chooses a
family threshold, component count, nuisance model, or transfer permutation
count. `report.html` includes an analysis-status table so “not supplied” cannot
be mistaken for “tested and nonsignificant.” The [outputs page](../outputs)
documents every file and column.

## Adapting the workflow

For mouse data, use `org.Mm.eg.db`, `species = "Mus musculus"`, and
`db_species = "MM"`. For human data, use `org.Hs.eg.db`,
`species = "Homo sapiens"`, and `db_species = "HS"`. Another organism uses its
`OrgDb` plus a compatible curated or custom vocabulary. Any categorical outcome
can be the phenotype by setting `phenotype_column` and `phenotype_reference` in
`chorale_control()`.

Always decide normalisation, identifiers, vocabulary, reference level,
covariates, local nuisance variables, exchangeability blocks, component counts,
and permutation resolution before reading the significant rows.
